Reformulating phylogenetic models to improve flexibility and speed

### 2018-05-08 main.tex
first pass of the new intro, it is at the right length and a new story

Ever-increasing capabilities of data collection techniques (i.e. genetic sequencing, geographic information systems monitoring, and etc) and sampling efforts from field biologist are making large volumes of ecological/biological data available over an ever-wider range of species and also higher resolution (i.e. measurable phenotypic traits to genetic sequence data). 
As a result, researchers are interested in exploring more complicated systems and fitting more models to data or synthetic databases of species traits.
One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis; that is, analyzing relationships (signals) among species when estimating the effects of species traits within a community. 
Although there are statistical tools available and widely used for this type of problem, these tools are either inflexible, or computationally inefficient and unpractical, or both when fitting large volumes of data or complex models.
In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic signal via mixed effect models, highlighting various complexities of phylogenetic signals through random effects.

In the last few decides, researcher have constantly develop new phylogenetic comparative methods and have shown importance of including phylogenetic signal in these type of problems \cite{felsenstein1985phylogenies, li2017statistical}. 
The core of the existing techniques are based on different modelling assumptions of the evolutionary process using various parametrizations of phylogenetic correlations (PC) of taxa; that is matrices inferred from the topology of evolutionary trees by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods. 
For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods assuming Brownian motion (pure genetic draft) process. 
Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, simply looking at phylogenetic signal among species via traits. \citep{li2017canfun, freckleton2002phylogenetic}.

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches. 
More specifically, we will show our new formulation is more flexible and can be easily extended along with the features of mixed model such as phylogenetically correlated random slopes (i.e. the relationship among taxa via changes of predictors/covariates on traits), random interactions \ml{why not? we are already using this for the alternative formulation of nested RE} and nested random effect models.
We will compare our technique with existing methods to show robustness and efficiency via simulations and data.
We will end off by discussing opportunities and practicalities of our method and reevaluate the state of art for this area of research.

>>>>

Ever-increasing capabilities of data collection techniques (i.e. genetic sequencing, geographic information systems monitoring, and etc) and sampling efforts from field biologist are making large volumes of ecological/biological data available over an ever-wider range of species and also higher resolution (i.e. measurable phenotypic traits to genetic sequence data). 
As a result, researchers are interested in exploring ways to study more complicated systems and fitting more models to data or synthetic databases of species traits.
One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis using phylogenetic (eco-evolutionary) regression; that is, analyzing evolutionary relationships (signals) among species between species traits and environmental variables when estimating the effects of traits within a community.
Although there are sophisticated statistical tools specially designed for this type of problem, the current state of the art of these tools are either inflexible, or computationally inefficient and unpractical when fitting large volumes of data.
In particular, modelling and understanding relationships between response and multiple predictor variables, and accounting for different levels of variation (errors) to complicated systems with large data is always challenging both biologically and statistically, and adding phylogenetic relationships will further increases the difficulties. 
In such case, researchers often try to find ways to simplify or revert to simpler models (i.e. model reduction or ignore phylogenetic relationships, or both) for practical uses \cite{bunnefeld2012island, ord2010adaptation}, which can lead to both statistical and biological issues \cite{felsenstein1985phylogenies, li2017statistical}. 
In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relationships through random effects.
We will use this modelling framework to present and discuss importance of different levels of variations and their considerations and interpretations in phylogenetic comparative analysis.

There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
First, a method for linking phylogenetic relationships among species when estimating the effects of traits. 
The standard/classic phylogenetic regression is a correlated-residual model using phylogenetically independent contrasts (PICs), where the residuals evolve as a Brownian motion process \citep{felsenstein1985phylogenies}; in other words, the variation of residuals are phylogenetically correlated.
Many of the more recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} -- are built upon Felsentein's PICs model by making additional sophisticated assumptions allowing extra parameters to correct for possible bias where separates/partitions the phylogenetically correlated residual variation into phylogenetically uncorrelated residual variation (observation error or tip variation) and phylogenetic signal (biological/evolutionary process error) \cite{hansen2012interpreting}.
However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error and residual variation is associated with observation level variation (the standard statistical definition). 
Despite the addition of multiple observations being straightforward statistically, it is rarely seen in the literature as existing platforms are unable to fit multiple observations, or unbalanced design data, or both.

The second challenge is flexibly-incorportating modelling phylogenetic relationships to other common statistical features with many predictor variables in phylogenetic regression (i.e. phylogenetically correlated slopes, interactions, nested designs and etc.)
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species, \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance. 
The tools available for extending phylogenetic relationships to common statistical regression features are very limited, inflexible and computationally inefficient; thus, sophisticated biologist would turn to more flexible settings (e.g. Bayesian models).

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches. 
More specifically, we will show our new formulation is more flexible and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetically phylogenetic relationship on other predictor variables with random slopes (i.e. phylogenetic signal in response to change in the predictor variable), random interactions \ml{why not? we are already using this for the alternative formulation of nested RE} and nested random effect models.
We will compare our technique with existing methods to show robustness and efficiency via simulations and data.
We will end off by discussing opportunities and practicalities of our method and reevaluate the state of art for this area of research.

### 2018-05-08 phyloglmm_ms.tex
pushing to bb
temp pushing bits and pieces
first pass of the new intro, it is at the right length and a new story

\documentclass[12pt]{article}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage{tabularx}
% use \xspace to allow for space after a macro as necessary
\usepackage{xspace}
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage{pdfpages}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage[sort]{natbib}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage{hyperref}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{Reformulating phylogenetic models to improve speed and flexibility}
\author{Michael Li and Ben Bolker}
\date{\today}

@@

\keywords{phyloglmm, species--branch matrix, phylogenetic correlation}

>>>>

\documentclass[12pt]{article}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[]{soul}
\usepackage{tabularx}
% use \xspace to allow for space after a macro as necessary
\usepackage{xspace}
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage{pdfpages}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage[sort]{natbib}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage{hyperref}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{Reformulating phylogenetic models to improve flexibility and speed}
\author{Michael Li and Ben Bolker}
\date{\today}

@@

\keywords{phyloglmm, species--branch matrix, phylogenetic correlation, phylogenetic comparative methods}
